\Large\bf FlexibleSUSY --- A spectrum generator generator for supersymmetric models

### 2014-06-03 doc/paper.tex
Improvements suggested by Dominik

\section{Overview of the Program}
\label{Sec:Program}
\fs is a \mathematica package designed to create a fast and easily
adaptable spectrum generator in C++ for any SUSY model and some
non-supersymmetric models.  The user specifies the model by giving the
superfield content, superpotential, gauge symmetries and mass mixings
in the \sarah model files, examples of which can be found in the
\code{Models} directory of \sarah and the \code{sarah} directory of
FlexibleSUSY.  The user then also specifies the boundary conditions on
the model parameters in a separate \code{FlexibleSUSY.m} file,
examples of which can be found in the \code{model_files} directory. A
detailed guide on how to write the FlexibleSUSY.m model is given in
section \ref{Sec:modfile}.

\fs makes use of the existing \sarah package
\cite{Staub:2010ty,Staub:2009bi,Staub:2010jh,Staub:2012pb,Staub:2013tta}
to obtain tree level expressions for the mass matrices and electroweak
symmetry breaking conditions, one-loop self energies, one-loop
tadpoles corrections and two-loop renormalisation group
equations(RGEs) for the model specified in the model files. Additional
corrections which have been calculated elsewhere, such as two-loop
corrections to the Higgs masses\footnote{By default FlexibleSUSY has
  two-loop corrections to the Higgs masses for the
  MSSM\cite{Degrassi:2001yf,Brignole:2001jy,Dedes:2002dy,Brignole:2002bz,Dedes:2003km}
  and NMSSM \cite{Degrassi:2009yq} in FORTRAN files supplied by Pietro
  Slavich. These are the same corrections which are implemented in
  many of the public spectrum generators.} may be added in file
specified by the user.

\fs then converts these algerbraic expressions from \sarah into C++
code using the C++11 standard. \fs uses some parts of
SOFTSUSY\cite{Allanach:2001kg}, the very fast Eigen library
\cite{eigen} augmented by LAPACK,
as well as the gnu scientific library and the boost
library to create numerical routines which solve the RGEs and EWSB
conditions.  The main solver routine of \fs then calls these routines
to find simultaneous solutions for the EWSB, low energy data, and
other, user supplied, boundary conditions (e.g.\ the usual mSUGRA
universality conditions) and calculates the mass spectrum.

The modular structure of the generated code allows for any individual
components to be replaced with an alternative if available.  Indeed
\fs has been intentionally designed to grow as alternative solvers and
calculators are added and a subsequent release with an alternative
lattice solver is already planned.  In addition if one already has an
existing code for some other purpose and simply wishes to improve upon
this by adding, e.g., new RGEs or self energies the modular framework
makes this straightforward by adding or interfacing with the
appropriate routine.

\subsection{Design goals}
\label{sec:design goals}

FlexibleSUSY is designed with the following points in mind:

@@

\paragraph{Modularity}
The large variety of supersymmetric models makes it likely that the
user wants to modify the generated spectrum generator source code.
\fs uses C++ object orientation features to modularize the source code
in order to make it easy for the user to modify, reuse and extend the
spectrum generator.  An important application of this concept are the
boundary conditions: The RG solver provides a common \code{Constraint}
interface, see \figref{fig:schematic-two-scale-constraint-interface}.
%
\begin{figure}
  \centering
  \tikzumlset{fill class=white}
  \begin{tikzpicture}
    \umlclass[x=0, y=0]{RGFlow<Two\_scale>}{}{
      + solve()
    }
    \umlclass[x=6, y=0, type=abstract]{Constraint<Two\_scale>}{}{
      + \umlvirt{apply()}\\
      + \umlvirt{get\_scale() : double}
    }
    \umlclass[x=6, y=-3]{MyConstraint}{}{
      + apply()\\
      + get\_scale() : double
    }
    \umldep{RGFlow<Two\_scale>}{Constraint<Two\_scale>}
    \umlinherit{MyConstraint}{Constraint<Two\_scale>}
  \end{tikzpicture}
  \caption{Schematic two-scale constraint interface}
  \label{fig:schematic-two-scale-constraint-interface}
\end{figure}
%
To create a custom constraint \code{MyConstraint}, the user implements
the interface by inheriting from \code{Constraint} and writing the
\code{apply()} and \code{get_scale()} methods (see
\secref{sec:boundary-conditions} for more details).  The so created
constraint class can then be passed to the RG solver to impose the
user-defined boundary condition at the chosen scale.

@@

\subsection{Adapting C++ code}

@@

>>>>

\section{Overview of the Program}
\label{Sec:Program}

To study the properties of SUSY models programs are needed which
numerically calculate the pole masses and couplings of the SUSY
particles given a set of theory input parameters.  The output of these
so-called spectrum generators can be transferred to programs which
calculate further observables such as branching rations or the dark
matter relic density.

In order to create a spectrum generator the SUSY model must be defined
by specifying the gauge group, the field content and mixings as well
as the superpotential and the soft-breaking terms.  From this
information the renormalization group equations, mass matrices,
self-energies, tadpole diagrams and electroweak symmetry breaking
conditions have to be derived.  These expressions must then be
combined in a computer program to allow for a numeric calculation of
the mass spectrum.  In addition most SUSY models require boundary
conditions for the model parameters at a low and a high scale.  For
example in the CMSSM mSUGRA boundary conditions for the soft-breaking
parameters are imposed at the gauge coupling unification scale.
Furthemore, at the $Z$ mass scale the CMSSM is matched to the Standard
Model, which implies conditions for the gauge and Yukawa couplings.
The so defined boundary value problem must be solved numerically until
a set of model parameters has been found consistent with all
user-supplied boundary conditions.

\fs is a \mathematica package designed to create a fast and easily
adaptable spectrum generator in C++ for any SUSY model and some
non-supersymmetric models.  The user specifies the model by giving the
superfield content, superpotential, gauge symmetries and mass mixings
in the \sarah model files.  Furthermore, the boundary conditions on
the model parameters must be specified in a separate
\code{FlexibleSUSY.m} file.
%
Based on this information \fs uses the existing \sarah package
\cite{Staub:2010ty,Staub:2009bi,Staub:2010jh,Staub:2012pb,Staub:2013tta}
to obtain tree level expressions for the mass matrices and electroweak
symmetry breaking conditions, one-loop self energies, one-loop
tadpoles corrections and two-loop renormalisation group equations
(RGEs) for the model.  Additional corrections which have been
calculated elsewhere, such as two-loop corrections to the Higgs
masses\footnote{By default FlexibleSUSY has two-loop corrections to
  the Higgs masses for the MSSM
  \cite{Degrassi:2001yf,Brignole:2001jy,Dedes:2002dy,Brignole:2002bz,Dedes:2003km}
  and NMSSM \cite{Degrassi:2009yq} in FORTRAN files supplied by Pietro
  Slavich. These are the same corrections which are implemented in
  many of the public spectrum generators.} may be added in file
specified by the user.
%
These algerbraic expressions are converted into C++ code and are put
into classes with well-defined interfaces to allow for easy exchange,
extension and reuse of the modules.  All of these classes are finally
combined to a complete spectrum generator, which solves the
user-defined boundary value problem.  For this task \fs uses some
parts of SOFTSUSY \cite{Allanach:2001kg}, the very fast Eigen library
\cite{eigen}, augmented by LAPACK, as well as the GNU scientific
library and the BOOST library to create numerical routines which solve
the RGEs and boundary conditions simultaneously.  If a solution has
been found the pole mass spectrum is eventually calculated using full
one-loop self-energies (and leading two-loop Higgs self-energy
contributions for the MSSM and NMSSM).

\subsection*{Design goals}

Since the calculation of the pole mass spectrum in a SUSY model is a
non-trivial task, \fs is designed with the following points in mind:

@@

\paragraph{Modularity}

The large variety of supersymmetric models makes it likely that the
user wants to modify the generated spectrum generator source code or
reuses componets in his own program. \fs uses C++ object orientation
features to modularize the source code in order to make it easy for
the user to modify, reuse, replace or extend the individual
components.  An important application of this concept are the boundary
conditions: The boundary conditions solver class provides a plugin
mechanism via a common \code{Constraint} interface, which allows a
user to exchange or add boundary conditions at any scale.
Alternatively, if one already has an existing code for some other
purpose and simply wishes to improve upon this by adding, e.g., new
RGEs or self-energies the modular framework makes this straightforward
by adding or interfacing with the appropriate routine.  A possible
application of reusing the \fs' generated classes would be to further
improve the precision of the CE$_6$SSM spectrum generator presended in
\cite{Athron:2009bs,Athron:2012pw} by adding further one-loop
self-energies and two-loop $\beta$-functions.

\paragraph{Alternative boundary condition problem solvers}

Furthermore, the standard algorithm which solves the user-defined
boundary value problem via a fixed-point iteration is not guaranteed
to converge in all regions of the model parameter space.  Therefore,
\fs has been intentionally designed to allow for alternative solvers
to search for solutions in such critical parameter regions.  A
subsequent release with an alternative lattice solver is already
planned.

\paragraph{Towers of effective theories}

In \fs the standard fixed-point iteration solver has been generalized
to handle towers of models (effective theories), which are matched at
intermediate scales.  An example of such a tower construction will be
given in section \ref{sec:adapting-cpp-code}, where a right-handed
neutrino is integrated out at the see-saw scale, between the SUSY and
the GUT scale.

@@

\subsection{Adapting C++ code}
\label{sec:adapting-cpp-code}

@@

%\cite{Athron:2009bs}
\bibitem{Athron:2009bs}
  P.~Athron, S.~F.~King, D.~J.~Miller, S.~Moretti and R.~Nevzorov,
  %``The Constrained Exceptional Supersymmetric Standard Model,''
  Phys.\ Rev.\ D {\bf 80} (2009) 035009
  [arXiv:0904.2169 [hep-ph]].
  %%CITATION = ARXIV:0904.2169;%%
  %45 citations counted in INSPIRE as of 03 Jun 2014

%\cite{Athron:2012pw}
\bibitem{Athron:2012pw}
  P.~Athron, D.~Stockinger and A.~Voigt,
  %``Threshold Corrections in the Exceptional Supersymmetric Standard Model,''
  Phys.\ Rev.\ D {\bf 86} (2012) 095012
  [arXiv:1209.1470 [hep-ph]].
  %%CITATION = ARXIV:1209.1470;%%
  %5 citations counted in INSPIRE as of 03 Jun 2014
